Effectful Semantics for Substructural SSA

### 2023-11-17 pldi24/paper.tex
Tap?

In a standard SSA-based compiler, to represent a function, we begin by
decomposing it into basic blocks, which can be viewed as \textit{label},
followed by a sequence of \textit{instructions}, each defining a
\textit{variable}, followed by a \textit{terminator}. A program written in this
manner being in "SSA form" simply means that every variable name is defined
exactly once. For example, the program in Figure \ref{fig:3-addr} has 3 basic
blocks, one of which is a distinguished \textit{entry block}, but is
\textit{not} in SSA form, since the variables \(m, n\) and \(i\) are each
modified through subsequent assignments (\(t\) is recomputed on every iteration
of the loop, but since it only has one definition, it \textit{is} in SSA form).
To convert this program to SSA form, we would traditionally use
\(\phi\)-functions, or, equivalently, annotate each basic block with arguments,
as in Figure \ref{fig:bb-arg}.

@@

For our type theoretic presentation of SSA, we will treat a very slight
generalization of the standard basic blocks with arguments syntax, in that we
will support \textit{nested scopes} and, in particular, \textit{if-statements}.
The program in Figure \ref{fig:bb-arg} translated to our syntax is given in
Figure \ref{fig:our-syn}. These do not introduce any new expressivity: an
if-statement can be viewed as syntactic sugar for a conditional branch to one of
two anonymous basic blocks, while nested scopes simply \textit{restrict} where
we can branch to. 
% In particular, in Section \ref{ssec:ssa-norm}, we will formalize
% interconversion with traditional SSA and show how it is semantically sound to
% simply merge all nested scopes. The point of the slightly modified syntax is
% simply to make some of our metatheoretic results more practical to state.

The formal grammar for our syntax is given in Figure \ref{fig:ssa-grammar} in
the appendix, while the typing judgements themselves are given in
\ref{fig:ssa-judgements}. We begin by giving a grammar for \textit{types} \(A,
B, C\), parametrized by a set of \textit{base types} \(X\); all syntactically
well-formed types are also semantically valid. We postulate the existence of a
\textit{unit type} \(\mb{1}\), a \textit{boolean type} \(\mb{2}\), and all
tensor products \(A \otimes B\); we write \(\otimes\) to underscore the fact
that types are allowed to be substructural, assigning every type a linearity
\(\ms{lin}(A)\) satisfying the constraints in Figure \ref{fig:ssa-structural}.
In particular, the rules \rle{unit-lin} and \rle{bool-lin} state that both the
unit and boolean types must be intuitionistic; the rule \rle{pair-lin} says that
the linearity of a pair is simply the intersection of the linearity of its
components. The rules \rle{nil-lin} and \rle{cons-lin} together say that the
linearity of a context is the intersection of the linearity of its variables,
where the linearity of a variable is the intersection of the linearity of its
type and its quantity annotation.

Our grammar proper splits program fragments into three main syntactic categories:
\begin{itemize}
  \item \textit{Expressions} \(a, b, c, d, e\), which are typed via the
  judgement \(\hasty{\Gamma}{p}{a}{A}{q}\), which says that the term \(a\)
  is of type \(A\) with \textit{centrality} \(p = 1\) if \textit{central}, or
  \(p = 0\) otherwise, and \textit{quantity} \(q \subseteq \{\ms{a}, \ms{r}\}\).
  So the centrality \(p\) indicates whether it is safe to substitute this
  term for a variable, and \(q\), plus the linearity of \(A\), control how
  many times that variable can occur: if \(\taff \subseteq q\) then the variable
  is called \textit{affine} or \textit{discardable} and it is permissible to
  leave it unused, whereas if \(\trel \subseteq q\) it is called
  \textit{relevant} or \textit{duplicable} and it is permissible to use multiple
  times. An term which is central, relevant, and affine is called
  \textit{pure}. The rules for this judgement are given in Figure
  \ref{fig:ssa-term-typing}.

@@

\item \textit{Control-flow graphs} or \textit{CFG}s \(L\), which are typed via
  the judgement \(\lhaslb{\ms{L}}{L}{\ms{K}}\), which says that \(L\) defines a
  control-flow graph with input labels \(\ms{L}\) and output labels \(\ms{K}\);
  recursive control flow is allowed. The rules for this judgement are given in
  Figure \ref{fig:ssa-block-typing}.
\end{itemize}
Here \textit{contexts} \(\Gamma\), consist of a list of variables \(x\) given a
type \(A\) annotated with a \textit{quantity} \(q \subseteq \tint\). A variable
\(x: A^q\) is called \textit{affine} if its type \(A\) is affine (i.e.
\(\tlin{A}{\taff}\)) \textit{and} \(\taff \subseteq q\), in which case it can be
freely ignored. Similarly, \(x\) is \textit{relevant} if its type \(A\) relevant
(i.e. \(\tlin{A}{\trel}\)) \textit{and} \(\trel \subseteq q\), in which case it
can be freely ignored. This is formalized in the structural judgement
\(\csplits{\Gamma}{\Delta}{\Xi}\), which says that \(\Gamma\) can be split into
subcontexts \(\Delta\) and \(\Xi\) (weakening is simply the special case
\(\cwk{\Gamma}{\Delta} \iff \csplits{\Gamma}{\Delta}{\cdot}\)), as per the
typing rules in Figure \ref{fig:ssa-structural}. In particular, the rule
\rle{split-nil} provides a base case by saying that the empty context can be
split into two empty contexts, while the rules \rle{split-left} and
\rle{split-right} send \(x: A^q\) to the left or right subcontext repsectively;
both these rules also allow us to reduce the quantity \(q\) to a quantity \(r
\subseteq q\), "forgetting" some of our capabilities. The rule \rle{split-dup}
allows the duplication of a variable which has a relevant type \(A\)
\textit{and} a relevant quantity \(q\), with the option to weaken the quantity
to \textit{different} quantities \(r\) and \(s\) on the left-hand side and
right-hand sides respectively. Finally, the rule \rle{split-aff} simply allows
us to ignore a variable \(x: A^q\) if \(A\) and \(q\) are affine.

@@

We can now get to the typing judgement for expressions in Figure
\ref{fig:ssa-term-typing}. The rule \rle{var} is essentially the standard rule
for variables, with the requirement that, other than \(x\), all variables in
\(\Gamma\) are affine. Note that the quantity \(q\) of the \textit{term}
\(x\) is must satisfy \(q \subseteq q'\), where \(x: A^{q'} \in \Gamma\); this
reflects that we're allowed to \textit{forget} capabilities, but not
\textit{add} them. The \rle{app} rule is also mostly standard: given an
instruction \(f\) which takes in an argument \(A\) and has a return type of
\(B\) and a \textit{central} term \(\hasty{\Gamma}{1}{a}{A}{q}\), we
conclude that \(\hasty{\Gamma}{p}{f\;a}{B}{q}\). We, of course, require that
\(f\) itself have centrality \(p\) and quantity \(q\), i.e. that \(f \in
\mc{I}_p^q(A, B)\). The \rle{pair} rule says that we can type a pair \((a, b)\)
in \(\Gamma\) if \(\csplits{\Gamma}{\Delta}{\Xi}\) where \(\Delta\) types \(a\)
and \(\Xi\) types \(b\); the quantity of a pair is simply the intersection of
the quantity of its components (this is expressed in the rule by allowing any
quantity which types \textit{both} components), and both components must be
central. Finally, we have the rules \rle{let} and \rle{let2}, corresponding to
let-bindings and destructuring let-bindings respectively. In both cases, we
split the context \(\Gamma\) into subcontext \(\Delta, \Xi\), one of which is
used to type a central subterm \(a\), which is bound to variables \(x\) or
destructured to variables \(x, y\) respectively, in a potentially non-central
term \(e\) which is otherwise typed in \(\Delta\).

>>>>

In a standard SSA-based compiler, to represent a function, we begin by
decomposing it into basic blocks, which can be viewed as \textit{label},
followed by a sequence of \textit{instructions}, each defining a
\textit{variable}, followed by a \textit{terminator}. A program written in this
manner being in "SSA form" simply means that every variable name is defined
exactly once. For example, the program in Figure \ref{fig:3-addr} has 3 basic
blocks, one of which is a distinguished \textit{entry block}, but is
\textit{not} in SSA form, since the variables \(m, n\) and \(i\) are each
modified through subsequent assignments .
% (\(t\) is recomputed on every iteration of the loop, but since it only has one
% definition, it \textit{is} in SSA form).
To convert this program to SSA form, we would traditionally use
\(\phi\)-functions, or, equivalently, annotate each basic block with arguments,
as in Figure \ref{fig:bb-arg}.

@@

For our type theoretic presentation of SSA, we will treat a very slight
generalization of the standard basic blocks with arguments syntax, in that we
will support \textit{nested scopes} and, in particular, \textit{if-statements}.
The program in Figure \ref{fig:bb-arg} translated to our syntax is given in
Figure \ref{fig:our-syn}. These do not introduce any new expressivity: an
if-statement can be viewed as syntactic sugar for a conditional branch to one of
two anonymous basic blocks, while nested scopes simply \textit{restrict}
branching. Our syntax is composed of three main syntactic categories:
\begin{itemize}
  \item \textit{Terms} \(a, b, c, d, e\), which are typed via the judgement
  \(\hasty{\Gamma}{p}{a}{A}{q}\), which says that the term \(a\) is of type
  \(A\) with \textit{centrality} \(p = 1\) if \textit{central}, or \(p = 0\)
  otherwise, and \textit{quantity} \(q \subseteq \{\ms{a}, \ms{r}\}\). So the
  centrality \(p\) indicates whether it is safe to substitute this term for a
  variable, and \(q\), plus the linearity of \(A\), control how many times that
  variable can occur: if \(\taff \subseteq q\) then the variable is called
  \textit{affine} or \textit{discardable} and it is permissible to leave it
  unused, whereas if \(\trel \subseteq q\) it is called \textit{relevant} or
  \textit{duplicable} and it is permissible to use multiple times. A term which
  is central, relevant, and affine is called \textit{pure}. The rules for this
  judgement are given in Figure \ref{fig:ssa-term-typing}.

@@

\item \textit{Control-flow graphs} or \textit{CFG}s \(L\), which are typed via
  the judgement \(\lhaslb{\ms{L}}{L}{\ms{K}}\), which says that \(L\) defines a
  control-flow graph with input labels \(\ms{L}\) and output labels \(\ms{K}\);
  recursive control flow is allowed. The rules for this judgement are given in
  Figure \ref{fig:ssa-block-typing}.
\end{itemize}
The formal grammar for our syntax is given in Figure \ref{fig:ssa-grammar} in
the appendix, while the typing judgements themselves are given in
\ref{fig:ssa-judgements}.

We postulate the existence of a set of \textit{base types} \(X\), as well as a
\textit{unit type} \(\mb{1}\) and \textit{boolean type} \(\mb{2}\). We define
the tensor product of types \(A \otimes B\); we write \(\otimes\) to underscore
the fact that types are allowed to be substructural, assigning every type a
linearity \(\ms{lin}(A)\) satisfying the constraints in Figure
\ref{fig:ssa-structural}. In particular, the rules \rle{unit-lin} and
\rle{bool-lin} state that both the unit and boolean types must be
intuitionistic; the rule \rle{pair-lin} says that the linearity of a pair is
simply the intersection of the linearity of its components. The rules
\rle{nil-lin} and \rle{cons-lin} together say that the linearity of a context is
the intersection of the linearity of its variables, where the linearity of a
variable is the intersection of the linearity of its type and its quantity
annotation.

Here \textit{contexts} \(\Gamma\), consist of a list of variables \(x\) given a
type \(A\) annotated with a \textit{quantity} \(q \subseteq \tint\). A variable
\(x: A^q\) is called \textit{affine} if its type \(A\) is affine (i.e.
\(\tlin{A}{\taff}\)) \textit{and} \(\taff \subseteq q\), in which case it can be
freely ignored. Similarly, \(x\) is \textit{relevant} if its type \(A\) relevant
(i.e. \(\tlin{A}{\trel}\)) \textit{and} \(\trel \subseteq q\), in which case it
can be freely ignored. This is formalized in the structural judgement
\(\csplits{\Gamma}{\Delta}{\Xi}\), which says that \(\Gamma\) can be split into
subcontexts \(\Delta\) and \(\Xi\) (weakening is simply the special case
\(\cwk{\Gamma}{\Delta} \iff \csplits{\Gamma}{\Delta}{\cdot}\)), as per the
typing rules in Figure \ref{fig:ssa-structural}. In particular, the rule
\rle{split-nil} provides a base case by saying that the empty context can be
split into two empty contexts, while the rules \rle{split-left} and
\rle{split-right} send \(x: A^q\) to the left or right subcontext repsectively;
both these rules also allow us to reduce the quantity \(q\) to a quantity \(r
\subseteq q\), "forgetting" some of our capabilities. The rule \rle{split-dup}
allows the duplication of a variable which has a relevant type \(A\)
\textit{and} a relevant quantity \(q\), with the option to weaken the quantity
to \textit{different} quantities \(r\) and \(s\) on the left-hand side and
right-hand sides respectively. Finally, the rule \rle{split-aff} simply allows
us to ignore a variable \(x: A^q\) if \(A\) and \(q\) are affine.

@@

We can now get to the typing judgement for expressions in Figure
\ref{fig:ssa-term-typing}. The rule \rle{var} states that \(x: A^q\) is
well-typed in \(\Gamma\) if all other variables are affine (i.e., can be
dropped) and \(q \subseteq q'\), where \(x: A^{q'} \in \Gamma\); this reflects
that we're allowed to \textit{forget} capabilities, but not \textit{add} them.
The \rle{app} rule states that, given an instruction \(f\) which takes in an
argument \(A\) and has a return type of \(B\) and a \textit{central} term
\(\hasty{\Gamma}{1}{a}{A}{q}\), we can conclude that
\(\hasty{\Gamma}{p}{f\;a}{B}{q}\). We, of course, require that \(f\) itself have
centrality \(p\) and quantity \(q\), i.e. that \(f \in \mc{I}_p^q(A, B)\). The
\rle{pair} rule says that we can type a pair \((a, b)\) in \(\Gamma\) if
\(\csplits{\Gamma}{\Delta}{\Xi}\) where \(\Delta\) types \(a\) and \(\Xi\) types
\(b\); the quantity of a pair is simply the intersection of the quantity of its
components (this is expressed in the rule by allowing any quantity which types
\textit{both} components), and both components must be central. Finally, we have
the rules \rle{let} and \rle{let2}, correspond to let-bindings and destructuring
let-bindings respectively. In both cases, we split the context \(\Gamma\) into
subcontext \(\Delta, \Xi\), one of which is used to type a central subterm
\(a\), which is bound to variables \(x\) or destructured to variables \(x, y\)
respectively, in a potentially non-central term \(e\) which is otherwise typed
in \(\Delta\).
